---
title: "Interactive overlays in Shiny"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Interactive overlays in Shiny}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`overshiny` provides draggable and resizable rectangular elements that
overlay plots in Shiny apps. This may be useful in applications where users
need to define regions on the plot for further input or processing.

Let's take a look at a simple user interface that includes two 
`overlayToken()`s, which are small labels that can be dragged onto the plot to
create new overlays, and an `overlayPlotOutput()`, which is a plot where the 
overlays will appear:

```{r, eval = FALSE}
library(shiny)
library(ggplot2)
library(overshiny)

# --- User interface ---
ui <- fluidPage(
    titlePanel("Overlay demo"),

    sidebarLayout(
        sidebarPanel(
            # Control whether overlays are displayed and whether they alter the plot
            checkboxInput("show_overlays", "Show overlays", value = TRUE),
            checkboxInput("enable_logic", "Enable overlay logic", value = TRUE),
            tags$hr(),

            # Select date range for the plot
            dateRangeInput("date_range", "Date range", start = "2025-01-01", end = "2025-12-31"),
            tags$hr(),

            # Overlay controls: tokens that can be dragged onto the plot
            h5("Drag tokens below onto the plot:"),
            overlayToken("grow", "Grow"),
            overlayToken("shrink", "Shrink")
        ),

        mainPanel(
            # Main plot with support for overlays
            overlayPlotOutput("plot", width = "100%", height = 300)
        )
    )
)
```

This sets up a sidebar layout, with controls on the left (including the overlay
tokens) and a display area on the right, which includes the plot the overlays
will be used with.

Now let's put together our server function. We start by setting up the overlays:

```{r, eval = FALSE}
# --- App logic ---
server <- function(input, output, session)
{
    # --- OVERLAY SETUP ---

    # Initialise 8 draggable/resizable overlays
    ov <- overlayServer("plot", 8, width = 56, # 56 days = 8 weeks default width
        data = list(strength = 50), snap = snapGrid())

    # Toggle overlay visibility based on checkbox
    observe({
        ov$show <- isTRUE(input$show_overlays)
    })
    
```

The call to `overlayServer()` initializes (up to) 8 overlays that we can use. 
It also sets the default width of new overlays to 56, which is in plot 
coordinates. We'll be plotting a time series, so this means 56 days (8 weeks). 
It provides the argument `data`, which is a list of additional attributes to be
associated with each overlay. Here we're specifying that each overlay will have 
an associated `strength` attribute, which we'll use to determine how much each
overlay affects the output. And finally, we use `snap = snapGrid()` to specify
a snapping function; the default parameters for `snapGrid()` ensure that each
overlay's position and width is snapped to the nearest whole number.

Then, we start with some of the reactive logic of the overlays. We have a 
checkbox in our UI to control whether the overlays are shown or not, and the
call to `observe()` makes the overlays show or hide based on the value of
this checkbox.

Continuing on:

```{r, eval = FALSE}
    # --- OVERLAY DROPDOWN MENU ---

    # Render dropdown menu when an overlay is being edited
    output$plot_menu <- renderUI({
        i <- req(ov$editing)  # Current overlay being edited
        fmt <- function(t) format(as.Date(t, origin = "1970-01-01"), "%b %d")

        list(
            div(paste(fmt(ov$cx0[i]), "–", fmt(ov$cx1[i]))),
            selectInput("plot_label", NULL, choices = c("Grow", "Shrink"), selected = ov$label[i]),
            sliderInput("plot_strength", "Strength", min = 0, max = 100, value = ov$data$strength[i]),
            dateInput("plot_cx", "Start date", value = ov$cx0[i]),
            sliderInput("plot_cw", "Duration", min = 1, max = floor(ov$bound_cw), value = ov$cx1[i] - ov$cx0[i])
        )
    })

```

Each overlay automatically has a dropdown menu for adjusting settings for the
overlay. By default, this only includes a "remove" button that can be used to 
remove the overlay. But we can add extra elements to these menus by using
`renderUI()`.

Since we created an `overlayPlotOutput()` with the output ID `"plot"`, 
`overshiny` has also created a UI output slot named `"plot_menu"` which is used 
to add extra elements to each overlay's dropdown menu. For our purposes, we'll
include a `div()` element which shows the date range for the overlay, a 
`selectInput()` to choose between "Grow" and "Shrink" type overlays, and a 
`sliderInput()` to choose the percentage "strength" associated with the overlay.
We also allow the user to manually enter the start date of each overlay 
(`"plot_cx"`) and the width of each overlay (`"plot_cw"`). Here, `"plot_label"`,
`"plot_cx"`, and `"plot_cw"` will automatically set the label, position, and
width of each overlay because `"label"`, `"cx"`, and `"cw"` are interpreted
specially by `overshiny`. `"strength"` doesn't have any special interpretation
so it will be applied to `ov$data$strength`. See the documentation for
`overlayServer()` for more details.

The line `i <- req(ov$editing)` just gets the index (1 to 8) of the current
overlay being edited. The call to `req()` ensures that the rest of the code
in the `renderUI()` call won't be run unless there is an overlay currently
being edited via its dropdown menu.

Note that above, each overlay has the same elements in its dropdown menu, but
we could choose to return different contents for the dropdown menu depending on 
which overlay is being edited.

Now let's make some data to plot based on the overlays and their properties:

```{r, eval = FALSE}
    # --- DATA PROCESSING BASED ON OVERLAY POSITION ---

    # Reactive dataset: oscillating signal modified by active overlays
    data <- reactive({
        date_seq <- seq(input$date_range[1], input$date_range[2], by = "1 day")
        y <- 1 + 0.5 * sin(as.numeric(date_seq) / 58)  # oscillating signal

        # Modify signal according to active overlays if logic is enabled
        if (isTRUE(input$enable_logic)) {
            for (i in which(ov$active)) {
                start <- as.Date(ov$cx0[i], origin = "1970-01-01")
                end <- as.Date(ov$cx1[i], origin = "1970-01-01")
                in_range <- date_seq >= start & date_seq <= end
                factor <- ov$data$strength[i] / 100
                y[in_range] <- y[in_range] * if (ov$label[i] == "Grow") (1 + factor) else (1 - factor)
            }
        }

        data.frame(date = date_seq, y = y)
    })

```

Above, we create a `reactive()` data.frame. We set up a sinusoidally-varying
time series, then (if the "Enable overlay logic" checkbox is checked) we either
"grow" or "shrink" this time series where it overlaps with each active overlay.
We're using the `ov` object returned by `overlayServer()` to do this.

Finally, we render the time series:

```{r, eval = FALSE}
    # --- RENDERING OF DATA ---

    # Render plot and align overlays to current axis limits
    output$plot <- renderPlot({
        plot <- ggplot(data()) +
            geom_line(aes(x = date, y = y)) +
            ylim(0, 3) +
            labs(x = NULL, y = "Signal")

        overlayBounds(ov, plot, xlim = c(input$date_range), ylim = c(0, NA))
    })
}
```

This just creates a `ggplot()` plot of the time series, and includes a call to
`overlayBounds()` at the end of the `renderPlot()` expression block to ensure
the overlays are aligned properly. `overlayBounds()` itself returns the plot so
this also returns our plot object to Shiny to be plotted.

Now all that's left is to run the app:

```{r, eval = FALSE}
# --- Run app ---
if (interactive()) {
    shinyApp(ui, server)
}
```
