---
title: "Interactive overlays in Shiny"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Interactive overlays in Shiny}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`overshiny` provides draggable and resizable rectangular elements that
overlay plots in Shiny apps. This may be useful in applications where users
need to define regions on the plot for further input or processing.

Let's take a look at a simple user interface that includes two 
`overlayToken()`s, which are small labels that can be dragged onto the plot to
create new overlays, and an `overlayPlotOutput()`, which is a plot where the 
overlays will appear:

```{r, eval = FALSE}
library(shiny)
library(ggplot2)
library(overshiny)

# --- User interface ---
ui <- fluidPage(
    titlePanel("Overlay demo"),

    sidebarLayout(
        sidebarPanel(
            # Control whether overlays are displayed and whether they alter the plot
            checkboxInput("show_overlays", "Show overlays", value = TRUE),
            checkboxInput("enable_logic", "Enable overlay logic", value = TRUE),
            tags$hr(),

            # Select date range for the plot
            dateRangeInput("date_range", "Date range", start = "2025-01-01", end = "2025-12-31"),
            tags$hr(),

            # Overlay controls: tokens that can be dragged onto the plot
            h5("Drag tokens below onto the plot:"),
            overlayToken("grow", "Grow"),
            overlayToken("shrink", "Shrink")
        ),

        mainPanel(
            # Main plot with support for overlays
            overlayPlotOutput("display", width = "100%", height = 300)
        )
    )
)
```

This sets up a sidebar layout, with controls on the left (including the overlay
tokens) and a display area on the right, which includes the plot the overlays
will be used with.

Now let's put together our server function. We start by setting up the overlays:

```{r, eval = FALSE}
# --- App logic ---
server <- function(input, output, session)
{
    # --- OVERLAY SETUP ---

    # Initialise 8 draggable/resizable overlays
    ov <- overlayServer("display", 8, width = 56, # 56 days = 8 weeks default width
        data = list(strength = 50), snap = snapGrid(),
        heading = dateHeading("%b %e"), select = TRUE)

    # Toggle overlay visibility based on checkbox
    observe({
        ov$show <- isTRUE(input$show_overlays)
    })
    
```

The call to `overlayServer()` takes as its first argument the ID of the 
`overlayPlotOutput()` from the UI. Here, we initialize (up to) 8 overlays that 
we can use. We also set the default width of new overlays to 56, which is in 
plot coordinates. We'll be plotting a time series, so this means 56 days 
(8 weeks).

The `data` argument to `overlayServer()` is a list of additional attributes to 
be associated with each overlay. Here we're specifying that each overlay will 
have an associated `strength` attribute, which we'll use to determine how much 
each overlay affects the output. We also use `snap = snapGrid()` to specify a 
snapping function; the default parameters for `snapGrid()` ensure that each
overlay's position and width is snapped to the nearest whole number.

The next two arguments to `overlayServer()` relate to the overlay dropdown 
menus. Each overlay automatically has a dropdown menu for adjusting settings 
for the overlay. By default, this only includes a "remove" button that can be 
used to remove the overlay. But we can add additional components to the 
dropdown in the call to `overlayServer()`.

The `heading` argument to `overlayServer()` allows us to add a small heading to
the top of each overlay's dropdown menu; we are using the built-in 
`dateHeading()` function here to specify the format of the heading. In this 
case, `"%b %e"` translates to the abbreviated month name followed by the
day of the month. So, for example, if the overlay extends over the x-axis range
1st January to 28th February, the heading will show `"Jan 1 – Feb 28"`. 
Finally, the `select = TRUE` argument means that each dropdown menu will also 
allow us to change the type of the overlay, i.e. `"Grow"` or `"Shrink"`.

After the call to `overlayServer()`, we start with some of the reactive logic 
of the overlays. We have a checkbox in our UI to control whether the overlays 
are shown or not, and the call to `observe()` makes the overlays show or hide 
based on the value of this checkbox.

Continuing on:

```{r, eval = FALSE}
    # --- OVERLAY DROPDOWN MENU ---

    # Render dropdown menu when an overlay is being edited
    ov$menu <- function(ov, i) {
        list(
            sliderInput("display_strength", "Strength", min = 0, max = 100, value = ov$data$strength[i]),
            dateInput("display_cx", "Start date", value = ov$cx0[i]),
            sliderInput("display_cw", "Duration", min = 1, max = floor(ov$bound_cw), value = ov$cx1[i] - ov$cx0[i])
        )
    }

```

Here, we add some additional, custom components to the overlay dropdown menu
by assigning a function to the variable `ov$menu`. We can also pass this
function in as the `menu` argument to `overlayServer()`, but the two approaches
are equivalent here. 

For our purposes, we'll add a `sliderInput()` to choose the percentage 
"strength" associated with the overlay. We also allow the user to manually 
enter the start date of each overlay (`"display_cx"`) and the width of each 
overlay (`"display_cw"`). Here, `"display_cx"` and `"display_cw"` will 
automatically set the position and width of each overlay because `"cx"` and 
`"cw"` are interpreted specially by `overshiny`. `"strength"` doesn't have any 
special interpretation so it will be applied to `ov$data$strength`. Note that 
the IDs of all these UI widgets start with `"display_"` because we gave our
`overlayPlotOutput()` the ID `"display"` in the UI. See the documentation for 
`overlayServer()` for more details.

It's important here to set the "starting value" for each of the three custom
input widgets using values from the `ov` object. If we just supplied some 
"default" value for each of these, this value would reset each time we opened 
the dropdown menu.

In this example, each overlay has the same elements in its dropdown menu, but
we could choose to return different contents for the dropdown menu depending on 
which overlay `i` is being edited.

Now let's make some data to plot based on the overlays and their properties:

```{r, eval = FALSE}
    # --- DATA PROCESSING BASED ON OVERLAY POSITION ---

    # Reactive dataset: oscillating signal modified by active overlays
    data <- reactive({
        date_seq <- seq(input$date_range[1], input$date_range[2], by = "1 day")
        y <- 1 + 0.5 * sin(as.numeric(date_seq) / 58)  # oscillating signal

        # Modify signal according to active overlays if logic is enabled
        if (isTRUE(input$enable_logic)) {
            for (i in which(ov$active)) {
                start <- as.Date(ov$cx0[i], origin = "1970-01-01")
                end <- as.Date(ov$cx1[i], origin = "1970-01-01")
                in_range <- date_seq >= start & date_seq <= end
                factor <- ov$data$strength[i] / 100
                y[in_range] <- y[in_range] * if (ov$label[i] == "Grow") (1 + factor) else (1 - factor)
            }
        }

        data.frame(date = date_seq, y = y)
    })

```

Above, we create a `reactive()` data.frame. We set up a sinusoidally-varying
time series, then (if the "Enable overlay logic" checkbox is checked) we either
"grow" or "shrink" this time series where it overlaps with each active overlay.
We're using the `ov` object returned by `overlayServer()` to do this.

Finally, we render the time series:

```{r, eval = FALSE}
    # --- RENDERING OF DATA ---

    # Render plot and align overlays to current axis limits
    output$display <- renderPlot({
        plot <- ggplot(data()) +
            geom_line(aes(x = date, y = y)) +
            ylim(0, 3) +
            labs(x = NULL, y = "Signal")

        overlayBounds(ov, plot, xlim = c(input$date_range), ylim = c(0, NA))
    })
}
```

This just creates a `ggplot()` plot of the time series, and includes a call to
`overlayBounds()` at the end of the `renderPlot()` expression block to ensure
the overlays are aligned properly. `overlayBounds()` itself returns the plot so
this also returns our plot object to Shiny to be plotted.

Now all that's left is to run the app:

```{r, eval = FALSE}
# --- Run app ---
if (interactive()) {
    shinyApp(ui, server)
}
```
